Prof. N. Sri Namachchivaya's importance in allowing me to complete this thesis cannot be overestimated. I thank him sincerely for his patience teaching and advising me. Prof. Matthew West graciously agreed to step in as my adviser of record. The help he has provided goes well above the requirements of this official role and his interest in this research has been stimulating. Professors Martin Ostoja-Starzewski and Lee DeVille round out my examination committee. Both have helped me understand my own research much better and I feel fortunate to have met them. My daily interactions with other students in the Aerospace Engineering Nonlinear Systems Group have been vital. I hope to keep in touch and collaborate for a long time. When I started at the University of Illinois, I did not know anyone in Champaign-Urbana, now some of my closest friends are here. For numerous wonderful evenings together, I have them to thank. My family has consistently supported me; their requests to explain my research in terms they can understand have encouraged me more than they may realize.

Financial support has been provided by Canada's Natural Sciences and Engineering Research Council and the United States National Science Foundation.

% \begin{itemize}
% \item{Prof. N. Sri Namachchivaya}
% \item{Prof. Matthew West}
% \item{Examination committee members}
% \item{Students in the Aerospace Engineering Nonlinear Systems Group}
% \item{Friends at the University of Illinois and in Urbana-Champaign}
% \item{Family}
% \item{Financial support from NSERC and NSF}
% \end{itemize}
